---
title: "HW #3: Interaction Effects and Multi-Group Analyses"
author: "Daniel Lewis"
date: "December 25, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Set up working environment

Load packages

```{r message = F, warning = F}
# Tidyverse #
library(tidyverse)
library(readxl)
library(broom)
library(corrr)
library(magrittr)
library(wrapr)
library(viridis)
library(directlabels)

# Markdown #
library(rmarkdown)
library(tinytex)
library(knitr)
library(printr)
library(kableExtra)

# Other #
library(psych)
library(here)
```

# Coding for Multi-Group Analyses

Import data and change variable names.

```{r}
tbl.2 <- read_xlsx(here("data", "BEER.xlsx"))
colnames(tbl.2) <- c("bubb", 'beer', 'salt')
```

View data table

```{r tables-tbl.2}
ks <- function(t, c) {
  kable(t, caption = c, format = 'html', escape = F, digits = 3) %>%
  kable_styling(full_width = F, position = 'left', fixed_thead = T)
  }

ks(tbl.2[1:10,], 'Beer Data (first 10 observations)')
```

## Dummy Coding

Dummy code variables

```{r, results = "asis"}
tbl.2 <- tbl.2 %>%
  mutate(beer.f = factor(beer),
         salt.f = factor(salt))

# rename columns for each single-column matrix in 'a' with its name in a
reCol <- function(a) {
  n <- names(a)
  for (i in 1:length(a)) {
    colnames(a[[i]]) <- c(str_sub(n[i], end = 4L))
  }
  a
}

tbl.2 %>%
  select_if(is.factor) %>%
  map(contrasts) %>%
  reCol() %>%
  ks("Dummy Codes")
```

Analyze dummy-coded data

```{r, results = 'asis'}
fit.2.a <- lm(bubb ~ beer.f + salt.f +
                beer.f * salt.f, tbl.2)

fit.2.a %>%
  tidy() %>%
  ks('Regression: Interaction of Beer and Salt')

tbl.2.e <- tibble(beer.f = factor(c("guinness", "guinness",
                                    "heineken", "heineken")),
                  salt.f = factor(c("no", 'yes', 'no', 'yes')))

tbl.2.b <- predict(fit.2.a, newdata = tbl.2.e) %>%
  set_names(c("irish", "irish & salty",
              "dutch", "dutch & salty")) %>%
  enframe(name = "combo", value = 'bubbles')

tbl.2.b %>%
  ks('Predicted Number of Bubbles')

1:2 %>%
  map_dbl(~ tbl.2.b$bubbles[.*2] / tbl.2.b$bubbles[.]) %>%
  set_names(c('guinness','heineken')) %>%
  enframe("beer", "salt multiplier") %>%
  ks('Effect of Salt by Beer')

```

## Contrast Coding

Contrast code variables

```{r}
tbl.2 <- tbl.2 %>%
  mutate_if(~ is.factor(.) && contrasts(.)[1] != -.5, list(c = ~ C(., (contr.sum(2)/(-2))))) %>%
  rename_at(vars(contains('_c')),
            list(~ paste(gsub('.f_c', '', .), 'c', sep = '.')))

tbl.2 %>%
  select_at(vars(ends_with('.c'))) %>%
  map(contrasts) %>%
  reCol %>%
  ks('Contrast Code')
```

Analyze contrast-coded data

```{r}
fit.2.b <- lm(bubb ~ beer.c + salt.c +
                beer.c * salt.c, tbl.2)

fit.2.b %>%
  tidy() %>%
  ks('Regression: Interaction of Contrast-Coded Beer and Salt')

mean(tbl.2$bubb)

tbl.2.c <- tbl.2.e %>%
  mutate_all(~ C(., (contr.sum(2)/(-2)))) %>%
  rename_all(~ paste(gsub('.f', '', .), 'c', sep = '.'))

tbl.2.d <- predict(object = fit.2.b, newdata = tbl.2.c) %>%
  set_names(c("irish", "irish & salty",
              "dutch", "dutch & salty")) %>%
  enframe(name = "combo", value = 'bubbles')

tbl.2.d %>%
  ks('Predicted Number of Bubbles (Contrast Codes)')

1:2 %>%
  map_dbl(~ tbl.2.d$bubbles[.*2] / tbl.2.d$bubbles[.]) %>%
  set_names(c('guinness','heineken')) %>%
  enframe("beer", "salt multiplier") %>%
  ks('Effect of Salt by Beer (Contrast Codes)')
```

# Probing Interactions

Import data and change variable names

```{r}
tbl.3 <- read_xlsx(here('data', 'wfint.xlsx')) %>%
  rename_all(tolower)
```


View data

```{r}
ks(tbl.3[1:10,], 'Work and Family Data (First 10 Observations)')
```

## a. Actual Predictors

Analysis of 'actual' predictors, i.e., actual amount of control, relationship quality, and security at work and with the family

```{r}
tbl.3 %>%
  names() %>%
  grep('wa|fa', ., value = TRUE) %>%
  map(~ paste('anx ~', .) %>% as.formula()) %>%
  map_df(~ lm(., tbl.3) %>% tidy()) %>%
  ks('Regression Output for Anxiety and Actual Control, Relationship Quality, and Security')
```

## b. Moderation by Gender

Analysis of gender as a moderating variable of effects on anxiety. Continuing to use 'actual' predictors.

Significant interactions are automatically highlighted with green text.

```{r, results = 'asis'}
tbl.3.b <- tbl.3 %>%
  names() %>%
  grep('wa|fa', ., value = TRUE) %>%
  map(~ paste('anx ~', ., '+ gen +', ., ' * gen') %>% as.formula()) %>%
  map_df(~ lm(., tbl.3) %>% tidy())

tbl.3.b %>%
  mutate(estimate = cell_spec(round(estimate, 3), "html",
                              color = ifelse(grepl(':', term) &
                                               (p.value < .05),
                                             "green", "black"))) %>%
  ks('Regression Output for Moderation by Gender')
```

Compute the intercepts and slopes

```{r}
fun.3.b <- function(i, t) {
  v.e <- t[["estimate"]]
  v.t <- t[["term"]]
  s.m <- v.e[i]
  i.m <- v.e[i - 1]
  s.w <- s.m + v.e[i + 2]
  i.w <- i.m + v.e[i + 1]
  result <- list(term = v.t[[i]],
                 intercept0 = i.m, slope0 = s.m,
                 intercept1 = i.w, slope1 = s.w)
  result
}

tbl.3.b.2 <- seq(2, 6 * 4, 4) %>%
  map_df(~ fun.3.b(., tbl.3.b))
  
tbl.3.b.2 %>%
  kable(col.names = c("attribute",
                      "intercept", "slope",
                      "intercept", "slope"),
        digits = 3) %>%
  kable_styling("striped", position = "left", full_width = F) %>%
  add_header_above(c(" " = 1, "Men" = 2, "Women" = 2))
```

Plot the simple intercepts and slopes

```{r}
vec.3.b <- c('need', 'intercept', 'slope')

tbl.3.b.3 <- tbl.3.b.2 %>%
{list(men = select(., 1:3), women = select(., c(1, 4:5)))} %>%
  map(~ set_colnames(., vec.3.b)) %>%
  bind_rows() %>%
  add_column(gen = factor(c(rep("male", 6), rep('female', 6))))

tbl.3.b.3 %>%
  filter(grepl('fa', need)) %>%
  mutate(new = paste(need, ifelse(gen == 'male',
                                  '-m', '-f'), sep = '')) %>%
  ggplot(aes(x = 1, y = intercept,
             xend = 7, yend = intercept + slope * 7,
             color = gen)) +
  geom_segment() +
  geom_dl(aes(label = new, color = gen),
          method = list(dl.trans(x = x - 1.5), 'last.bumpup')) +
  scale_x_continuous(expand = expand_scale(.25)) +
  scale_y_continuous(expand = expand_scale(.1)) +
  labs(x = "predictor", y = "anxiety", title = "Plot of Simple Slopes for Family-Related Predictors") +
  scale_color_discrete(name = "gender")
```

## c. Moderation by Importance

Analysis of importance as a moderating variable of effects on anxiety.

Significant interactions are automatically highlighted with green text.

```{r}
tbl.3.c.a <- tbl.3 %>%
  names() %>%
  grep('wa|fa', ., value = TRUE) %>%
  map2(grep('wi|fi', names(tbl.3), value = TRUE),
       ~ paste('anx ~', .x, '+', .y, '+',
               .x, '*', .y) %>% as.formula()) %>%
  map_df(~ lm(., tbl.3) %>% tidy())

tbl.3.c.a %>%
  mutate(estimate = cell_spec(round(estimate, 3), "html",
                              color = ifelse(grepl(':', term) &
                                               (p.value < .05),
                                             "green", "black"))) %>%
  ks('Regression Output for Moderation by Importance')
```

Compute the intercepts and slopes

```{r}
fun.3.c <- function(i, t, data) {
  v.e <- t[["estimate"]]  # vector of coefficient estimates
  v.t <- t[["term"]]      # vector of terms
  
  m.t <- v.t[[i + 1]]     # name of moderator
  m <- data[[m.t]]        # vector of moderator data
  s.m <- sd(m)            # standard deviation of moderator
  x.m <- mean(m)          # mean of moderator
  hi <- x.m + s.m         # +1 SD level of moderator
  med <- x.m              # mean level of moderator
  lo <- x.m - s.m         # -1 SD level of moderator
  
  int <- v.e[i - 1]       # intercept estimate
  me <- v.e[i]            # main effect estimate
  mo <- v.e[i + 1]        # main effect of moderator
  pr <- v.e[i + 2]        # coeffient of product term estimate
  
  s.h <- me + pr * hi     # +1 SD: slope
  i.h <- int + mo * hi    # +1 SD: intercept
  s.m <- me + pr * med    # mean: slope
  i.m <- int + mo * med   # mean: int
  s.l <- me + pr * lo     # -1 SD: slope
  i.l <- int + mo * lo    # -1 SD: int
  
  list(term = v.t[[i]],   # result
       int.low = i.l, slo.low = s.l,
       int.med = i.m, slo.med = s.m,
       int.hi = i.h, slo.hi = s.h)
}

tbl.3.c.2 <- seq(2, 6 * 4, 4) %>%
  map_df(~ fun.3.c(., tbl.3.c.a, tbl.3))
  
headers <- c('Intercept', 'Slope')

tbl.3.c.2 %>%
  kable(col.names = c("Attribute", rep(headers, 3)), digits = 3) %>%
  kable_styling("striped", position = "left", full_width = F) %>%
  add_header_above(c(" " = 1, "Low (-1 SD)" = 2, "Medium (Mean)" = 2, 'High (+1 SD)' = 2))
```

Plot the simple intercepts and slopes

```{r}
vec.3.c <- c('need', 'intercept', 'slope')

tbl.3.c.3 <- tbl.3.c.2 %>%
{list(low = select(., 1:3),
      medium = select(., c(1, 4:5)),
      high = select(., c(1, 6:7)))} %>%
  map(~ set_colnames(., vec.3.b)) %>%
  bind_rows() %>%
  add_column(lvl = ordered(c(rep("low", 6),
                            rep('medium', 6),
                            rep('high', 6)),
                          levels = c('low', 'medium', 'high')))

tbl.3.c.3 %>%
  filter(grepl('facon|fasec', need)) %>%
  mutate(new = paste(need, ifelse(lvl == 'low', '-l',
                                  ifelse(lvl == 'medium', '-m',
                                         '-h')), sep = '')) %>%
  ggplot(aes(x = 1, y = intercept,
             xend = 7, yend = intercept + slope * 7,
             color = lvl)) +
  geom_segment() +
  geom_dl(aes(label = new, color = lvl),
          method = list(dl.trans(x = x - 1.5), 'last.bumpup')) +
  scale_x_continuous(expand = expand_scale(.25)) +
  scale_y_continuous(expand = expand_scale(.1)) +
  labs(x = "predictor", y = "anxiety", title = "Plot of Simple Slopes for Significant Interactions with Importance") +
  scale_color_discrete(name = "Level")

ggsave("imp_simple.png", last_plot(), width = 6, height = 4.5, dpi = 600)
```
